Classless Java

### 2015-11-23 paper/sections/Appendix.tex
Comments on the proof

\subsection{Proof of LEMMA 1}\label{subsec:proof1}
\textbf{LEMMA 1. }
Assume that
\begin{align*}
\II_0\ &=\ \mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC\\
[\![\II_0]\!]\ &=\ \emptyset\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \methods' \cC
\end{align*}
If $\II_0$ satisfies $\dom(\C_0)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)$, then $[\![\II_0]\!]$ satisfies $\dom([\![\II_0]\!])=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)\cup\dom(\methods')$.
\begin{proof}
It suffices to prove $\dom([\![\II_0]\!])=\dom(\C_0)\cup\dom(\methods')$.
\begin{itemize}
\item If $m\notin\dom(\methods')$, by the definition of $\mBody$, it is obtained from $\override$. The first argument of $\override$, namely $\methods\cup\methods'(m)$, is equal to $\methods(m)$, and the second argument is not changed as well, thus the result of $\override$ is not changed during translation. Hence $m\in\dom([\![\II_0]\!])$ iff $m\in\dom(\C_0)$.
\item If $m\in\dom(\methods')$, we are to prove $m\in\dom([\![\II_0]\!])$. So if $m$ is the $\QM{of}$ method generated by $\ofMethod$, its $\mBody$ value is well defined, by the rule $\override(\method,\none)=\method$. On the other hand, if $m$ is generated by $\otherMethod$, namely $m$ is a $\QM{with-}$ or setter method, we can also get its $\mBody$ value, since in the definition of $\otherMethod$, we can see $\isWith$ and $\isSetter$ ensure the compatible subtyping relationship. Hence $m\in\dom([\![\II_0]\!])$ as well.
\end{itemize}
\end{proof}

\subsection{Proof of LEMMA 2}\label{subsec:proof2}
\textbf{LEMMA 2. }
If $\II_0$ OK, then $[\![\II_0]\!]$ OK.
\begin{proof}
By the rule \rn{t-Intf} in Figure~\ref{ET}, we divide it into two parts.

@@

Since $\II_0$ OK, all the existing default and static methods are well typed in $[\![\II_0]\!]$, except for the new method \QM{of}. It suffices to prove that it still holds for $\ofMethod(\C_0)$.

\haoyuan{How to more explicitly say the default methods are ok in the new interface table, so that it is sufficient to only prove the generated ofMethod is ok?}

@@

\begin{itemize}
\item For field typing,
\[\Gamma(\QM_ m_i)=\C_i\subtype \C_i\]
\item For method typing of getters,
\[\Gamma, m_i:\C_i, \QM{this}:\C_0, \Gamma^{\mh_i}\vdash m_i\in \C_i=\C^{\mh_i}\]
\item For method typing of \QM{with-} methods, by \rn{t-StaticInvk}, \haoyuan{how to do typing for the generated method itself?}
\[\Gamma, \overline{m_i:\C_i}, \QM{this}:\C_0, \es_i:\Cs_i, \Gamma^{\mh_i}\vdash \C_0\QM{.of}\oR\es_i\cR\in \C_0=C^{\mh_i}\]
\item For method typing of setter methods, \haoyuan{what about the \QM{with} method?}
\[\Gamma, \QM{this}:\C_0, \Gamma^{\mh_i}\vdash\QM{this}\in \C_0=\C^{\mh_i}\]
\item To prove that $\forall i$, $\mh_i\subtype\mBody(m^{\mh_i}, \C_0)$,
 \begin{itemize}
 \item For getters, \[\C_i\ m_i\oR\cR\QM; \in \fieldsFunc(\C_0) \Rightarrow \C_i\ m_i\oR\cR\QM; \in \dom(\C_0)\]
 \item For \QM{with-} methods,
  \[\begin{array}{l}
  \mBody(\QM{with#}\m_i,\C_0) \mbox{ is of form }\mh\QM; \\ \valid(\C_0)
  \end{array}\Rightarrow \isWith\Rightarrow \C_0\subtype\mBody(m^{\mh_i}, \C_0)\]
 \item For setters,
  \[\begin{array}{l}
  \mBody(\m_i,\C_0) \mbox{ is of form }\mh\QM; \\ \valid(\C_0)
  \end{array}\Rightarrow \isSetter\Rightarrow \C_0\subtype\mBody(m^{\mh_i}, \C_0)\]
 \end{itemize}
\item To prove that such a created object indeed implements all the abstract methods in $\dom(\C_0)$, we simply refer to $\valid(\C_0)$, since it guarantees each abstract method $\method$ to satisfy $\isField$, $\isWith$ or $\isSetter$. But that object includes all implementations for those cases, hence it is of type $\C_0$ by \rn{t-Obj}.
\end{itemize}

\noindent\textbf{Part II.} Next we check that in $[\![\II_0]\!]$, $$\dom([\![\II_0]\!])=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)\cup\dom(\methods')$$
And actually it is guaranteed by \textbf{LEMMA 1}.

@@

\subsection{Proof of THEOREM}\label{subsec:proof3}
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK, then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ either $\II$ OK or $\II$ is a subtype of $\II_0$.
\begin{proof}
\textbf{LEMMA 2} already proves that $[\![\II_0]\!]$ is OK. On the other hand, if some $\II$ is not a subtype of $\II_0$, the generated code in translation has no way to affect the domain of $\II$, which finishes our proof.
\end{proof}

>>>>

\subsection{Proof of LEMMA 1}\label{subsec:proof1}
\marco{this seams to be just a sublemma for Lemma2, I think you can mention it in the appendix only.}
\textbf{LEMMA 1. }
Assume that
\begin{align*}
\II_0\ &=\ \mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC\\
[\![\II_0]\!]\ &=\ \emptyset\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \methods' \cC
\end{align*}
If $\II_0$ satisfies $\dom(\C_0)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)$, then $[\![\II_0]\!]$ satisfies $\dom([\![\II_0]\!])=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)\cup\dom(\methods')$.
\begin{proof}
It suffices to prove $\dom([\![\II_0]\!])=\dom(\C_0)\cup\dom(\methods')$.
\begin{itemize}
\item If $m\notin\dom(\methods')$, by the definition of $\mBody$, it is obtained from $\override$. The first argument of $\override$, namely $\methods\cup\methods'(m)$, is equal to $\methods(m)$, and the second argument is not changed as well, thus the result of $\override$ is not changed during translation. Hence $m\in\dom([\![\II_0]\!])$ iff $m\in\dom(\C_0)$.
\item If $m\in\dom(\methods')$, we are to prove $m\in\dom([\![\II_0]\!])$. So if $m$ is the $\QM{of}$ method generated by $\ofMethod$, its $\mBody$ value is well defined, by the rule $\override(\method,\none)=\method$.\marco{need to add here that we know of is not in dom(..)} On the other hand, if $m$ is generated by $\otherMethod$, namely $m$ is a $\QM{with-}$ or setter method,
 we can apply the last clause of override, since in the definition of $\otherMethod$, we can see $\isWith$ and $\isSetter$ ensure the compatible subtyping relationship. Hence $m\in\dom([\![\II_0]\!])$ as well.
\end{itemize}
\end{proof}

\marco{ we need to state (may be as a new lemma):
forall expressions e, if under interface table I0,I1..In
$\Gamma\vdash e\in\C$, then  under interface table [[I0]],I1..In
$\Gamma\vdash e\in\_<:\C$
ProofScatch:
(1) observing otherMethods(), we see that [I] is structurally a subtype of I, thus all the expressions typed in an interface table with I, are well typed in an interface table with [I]. 
The only interesting case is method call (normal, super and static)
consider $\Gamma\vdash e_0.m(e_1..e_n)\in\C$, with $e_0$ of type I0,
if it is well typed in the first table, then in the other table it is still well typed:
the method still exists and its argument types are the same type.
The expressions 1..n may be now typed in a subtype, but this is ok since the premise of the rule ask for $\_<:\C$
}
\subsection{Proof of LEMMA 2}\label{subsec:proof2}
\textbf{LEMMA 2. }
If $\II_0$ OK, then $[\![\II_0]\!]$ OK.
\begin{proof}

By the rule \rn{t-Intf} in Figure~\ref{ET}, we divide it into two parts.

@@

Since $\II_0$ OK, \marco{by newLemma} all the existing default and static methods are well typed in $[\![\II_0]\!]$, except for the new method \QM{of}. It suffices to prove that it still holds for $\ofMethod(\C_0)$.

@@

\begin{itemize}
\item For field typing,
\[\Gamma(\QM_ m_i)=\C_i\subtype \C_i\]
\item For method typing of getters,
\[\Gamma, m_i:\C_i, \QM{this}:\C_0, \Gamma^{\mh_i}\vdash m_i\in \C_i\]
We know that $\C_i=\C^{\mh_i}$ since...\marco{finish}.
\item For method typing of \QM{with-} methods, by \rn{t-StaticInvk}, \haoyuan{how to do typing for the generated method itself?}
\marco{the next judgement is not ok, state something like:
We consider the with method for a arbitrary i field.
In this case the method body we have to type would be...
and here you can replace $e_i$ with the correct specialized expression}
\[\Gamma, \overline{m_i:\C_i}, \QM{this}:\C_0, \es_i:\Cs_i, \Gamma^{\mh_i}\vdash \C_0\QM{.of}\oR\es_i\cR\in \C_0\]
We know that $\C_0=C^{\mh_i}$ since ..\marco{ finish}
\item For method typing of setter methods, \haoyuan{what about the \QM{with} method?}\marco{we do not formalize it}
\[\Gamma, \QM{this}:\C_0, \Gamma^{\mh_i}\vdash\QM{this}\in \C_0=\C^{\mh_i}\]
\marco{no, you also need to show that the part x=var is correct, and again push I= out}
\marco{for under, now that the explanation have been improved, use the names of the new auxiliary functions}
\item To prove that $\forall i$, $\mh_i\subtype\mBody(m^{\mh_i}, \C_0)$,
 \begin{itemize}
 \item For getters, \[\C_i\ m_i\oR\cR\QM; \in \fieldsFunc(\C_0) \Rightarrow \C_i\ m_i\oR\cR\QM; \in \dom(\C_0)\]
 \item For \QM{with-} methods,\marco{ use if-then or other english forms instead of the logic arrow}
  \[\begin{array}{l}
  \mBody(\QM{with#}\m_i,\C_0) \mbox{ is of form }\mh\QM; \\ \valid(\C_0)
  \end{array}\Rightarrow \isWith\Rightarrow \C_0\subtype\mBody(m^{\mh_i}, \C_0)\]
 \item For setters,
  \[\begin{array}{l}
  \mBody(\m_i,\C_0) \mbox{ is of form }\mh\QM; \\ \valid(\C_0)
  \end{array}\Rightarrow \isSetter\Rightarrow \C_0\subtype\mBody(m^{\mh_i}, \C_0)\]
 \end{itemize}
\item To prove that such a created object indeed implements all the abstract methods in $\dom(\C_0)$, we simply refer to $\valid(\C_0)$, since it guarantees each abstract method $\method$ to satisfy $\isField$, $\isWith$ or $\isSetter$. But that object includes all implementations for those cases, hence it is of type $\C_0$ by \rn{t-Obj}.
\end{itemize}

\noindent\textbf{Part II.} Next we check that in $[\![\II_0]\!]$, $$\dom([\![\II_0]\!])=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)\cup\dom(\methods')$$
And actually it is guaranteed by \textbf{LEMMA 1}.
\marco{may be you may just delete lemma1 and put its proof here?}

@@

\subsection{Proof of THEOREM}\label{subsec:proof3}
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK, then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ either $\II$ OK or $\II$ is a subtype of $\II_0$.
\begin{proof}
\textbf{LEMMA 2} already proves that $[\![\II_0]\!]$ is OK. On the other hand, if some $\II$ is not a subtype of $\II_0$, 
\marco{by the newLemma, we know that all its methods are still well typed, and } the generated code in translation has no way to affect the domain of $\II$,\marco{ so rule t-interf can still be applied,} which finishes our proof.
\end{proof}

### 2015-11-23 paper/sections/Translation.tex
Comments on the proof

\subsection{Results}
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK, then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ either $\II$ OK or $\II$ is a subtype of $\II_0$.

@@

To understand this theorem statement, we need to understand three kind of guarantees that we can offer for safety:
\begin{itemize}
\item \textit{Self coherence}: the generated code itself is well-typed; type errors are not present in code the user have not written.
\item \textit{Client coherence}: all the client code (as for example method calls) that is well typed without the generation/instrumentation process is well typed also after the generation.
\item \textit{Heir coherence}: Interfaces (and in general classes) inheriting from instrumented code are well typed if they was well typed without the instrumentation.
\end{itemize}
The first two levels of safety are essential to our lightweight framework, nevertheless, the third one is not needed. In Java it implies not touching the annotated interface, and surely an alternative approach is to create a new interface and generate what we need. But since we are using Lombok for the translation, the transformation is applied to the annotated interface itself. But heir coherence is indeed an issue. See below how this is affected by our approach:

@@

Here is the client code for our example. With interface \Q@B@ annotated with \mixin, by the translation rule, it will generate a method ``\Q@B withX(int x);@'' inside, as an automatic return type refinement. But in that case, interface \Q@C@ doesn't type check.

Now we go back and read the statement of our theorem again. It illustrates that a program with the translation of \mixin applied still type checks if it did, except for the subtypes of the annotated interface (type safety of them is not guaranteed).

\haoyuan{Text needs improving.}

@@

\noindent\textbf{LEMMA 2. }
If $\II_0$ OK, then $[\![\II_0]\!]$ OK.

>>>>

\subsection{Results}
\textbf{THEOREM. }
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK,\\*
if  
$\valid(\C_0)$  and $\QM{of}\notin\dom(\C_0)$,
then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$\\*
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ either $\II$ OK or $\II$ is a subtype of $\II_0$.

@@

To understand this theorem statement, we need to understand three kind of guarantees that we can offer for safety:
\begin{itemize}
\item \textit{Self coherence}: the generated code itself is well-typed; type errors are not present in code the user have not wrote.
In our case it means that either \mixin{} produces in controlled way an understandable error, or the class can be successfully annotate and the generated code is well typed.
We guarantee \textit{Self coherence}.

\item \textit{Client coherence}: all the client code (as for example method calls) that is well typed without the generation/instrumentation process is well typed also after the generation.
That is, the annotation do not remove any functionality, is just
adding more behaviour.
We guarantee \textit{Client coherence}.

\item \textit{Heir coherence}: Interfaces (and in general classes) inheriting from instrumented code are well typed if they was well typed without the instrumentation.
This would require to not add any (default or abstract) method to the annotated interfaces, including type refinement.
We do not guarantee \textit{Heir coherence}.
Indeed consider the following example

@@

By the translation rule, \mixin would  generate in \Q@B@  a method ``\Q@B withX(int x);@''.
This would break \Q@C@.

\end{itemize}
%The first two levels of safety are essential to our lightweight framework, nevertheless, the third one is not needed. In Java it implies not touching the annotated interface, and surely an alternative approach is to create a new interface and generate what we need. But since we are using Lombok for the translation, the transformation is applied to the annotated interface itself. But heir coherence is indeed an issue. See below how this is affected by our approach:

%Now we go back and read the statement of our theorem again. It illustrates that a program with the translation of \mixin applied still type checks if it did, except for the subtypes of the annotated interface (type safety of them is not guaranteed).

%\haoyuan{Text needs improving.}

@@

\marco{ you need to use the Lemma/theorem macros}
\noindent\textbf{LEMMA 2. }
If $\II_0$ OK, then $[\![\II_0]\!]$ OK.
This is what we defined as self coherence before.
